DeepPicar: A Low-cost Deep Neural Network-based Autonomous Car

### 2018-04-19 paper/background.tex
revision

Accelerating AI workloads, especially inferencing
operations, has received a lot of attention from academia and industry
in recent years as applications of deep learning are broadening to include
areas of real-time embedded systems such as autonomous vehicles. These
efforts include the development of various heterogeneous architecture-based 
system-on-a-chips (SoCs) that may include multiple cores, GPU,
DSP, FPGA, and neural network optimized ASIC hardware~\cite{Jouppi2017}.
Consolidating multiple tasks on SoCs with a lot of shared hardware
resources while guaranteeing real-time performance is also an active
research area, which is orthogonal to improving raw
performance. Consolidation is necessary for efficiency, but unmanaged 
interference can nullify the benefits of consolidation~\cite{Kim2016}.
For these reasons, finding a good computing platform is a
non-trivial task, one that requires a deep understanding of the
workloads and the hardware platform being utilized.

The \emph{primary objectives of this study} are to understand (1) the
necessary computing performance for applying AI technology-based
robotics systems, and (2) the characteristics of the computing
platform to support such a workload.
To achieve these goals, we implement a low-cost autonomous car platform
as a case-study and systematically conduct experiments, which we will 
describe in the subsequent sections.

>>>>

Accelerating AI workloads, especially inferencing
operations, has received a lot of attention from academia and industry
in recent years as applications of deep learning are broadening to include
areas of real-time embedded systems such as autonomous vehicles. These
efforts include the development of various heterogeneous architecture-based 
system-on-a-chips (SoCs) that may include multiple cores, GPU,
DSP, FPGA, and neural network optimized ASIC hardware~\cite{Jouppi2017}.
Consolidating multiple tasks on SoCs with a lot of shared hardware
resources while guaranteeing real-time performance is also an active
research area, which is orthogonal to improving raw
performance. Consolidation is necessary for efficiency, but unmanaged 
interference can nullify the benefits of consolidation~\cite{Kim2016}.
%% For these reasons, finding a good computing platform is a
%% non-trivial task, one that requires a deep understanding of the
%% workloads and the hardware platform being utilized.

The \emph{primary objectives of this study} are (1) to understand the
necessary computing performance to realize deep neural network based
robotics systems, (2) to understand the characteristics of the
computing platform to support such workloads, and (3) to evaluate the
significance of contention in shared hardware resources and existing
mitigation techniques to address the contention problem.

To achieve these goals, we implement a low-cost autonomous car platform
as a case-study and systematically conduct experiments, which we will 
describe in the subsequent sections.

### 2018-04-19 paper/evaluation-cachepart.tex
revision

Cache partitioning is a well-known technique to improve isolation in
real-time systems by giving a dedicated cache space to each individaul
task or core. In this experiment, we use PALLOC~\cite{yun2014rtas}, a
page-coloring based cache partitioning solution for Linux, to
investigate the effect of cache partitioning in protecting DeepPicar's
DNN based controller.

\begin{figure}
  \centering
  \includegraphics[width=.5\textwidth]{figs/cache-mapping}
  \caption{Physical address mapping of L1/L2 caches of Broadcom
    BCM2837 processor in Raspberry Pi 3.}
  \label{fig:cache-mapping}
\end{figure}

Page coloring is a OS technique that controls the physical addresses
of memory pages. By allocating pages over non-overlapping sets of the
cache, the OS can effectively partition the cache.
Figure~\ref{fig:cache-mapping} shows the physical address
mapping of Raspberry Pi 3's BCM2837 processor, which has 32K private
L1 I\&D (4way) caches and a share 512KB L2 (16 way) cache. In order not
to partition the private L1 caches, we use the bit 13 and 14 for
coloring, which results in 4 usuable page colors.

@@

@@

In the first experiment, we investigate the cache space sensitivity of
DeepPicar's DNN-based control loop. Using PALLOC, we create 4
different cgroups which are configured to use 4, 3,
2, and 1 colors (100\%, 75\%, 50\% and 25\% of the L2 cache
space, respectively). We then execute the DNN control loop (inference)
on each cache partition (cgroup) and measure the average processing
time.

>>>>

Cache partitioning is a well-known technique to improve isolation in
multicore by giving a dedicated cache space to each individaul 
task or core. In this experiment, we use PALLOC~\cite{yun2014rtas}, a
page-coloring based kernel-level memory allocator for Linux.
Page coloring is a OS technique that controls the physical addresses
of memory pages. By allocating pages over non-overlapping sets of the
cache, the OS can effectively partition the cache.
Using PALLOC, we investigate the effect of cache partitioning in
protecting DeepPicar's DNN based controller.

\begin{figure} [h]
  \centering
  \includegraphics[width=.5\textwidth]{figs/cache-mapping}
  \caption{Physical address mapping of L1/L2 caches of Broadcom
    BCM2837 processor in Raspberry Pi 3.}
  \label{fig:cache-mapping}
\end{figure}

Figure~\ref{fig:cache-mapping} shows the physical address
mapping of Raspberry Pi 3's BCM2837 processor, which has 32K private
L1 I\&D (4way) caches and a share 512KB L2 (16 way) cache. In order not
to partition the private L1 caches, we use the bit 13 and 14 for
coloring, which results in 4 usuable page colors.

@@

In the first experiment, we investigate the cache space sensitivity of
DeepPicar's DNN-based control loop. Using PALLOC, we create 4
different cgroups which are configured to use 4, 3,
2, and 1 colors (100\%, 75\%, 50\% and 25\% of the L2 cache
space, respectively). We then execute the DNN control loop (inference)
on one core using a different cgroup cache partition, one at a time,
and measure the average processing time.

@@

### 2018-04-19 paper/evaluation.tex
..
revision

\subsection{Setup}

The Raspberry Pi 3 Model B platform used in DeepPicar equips a Broadcom
BCM2837 SoC, which has a quad-core ARM Cortex-A53 cluster,
running at up to 1.2GHz. Each core has a 32K private Instruction cache
and a 32K private Data cache, and all cores share a 512KB L2 cache.
The chip also includes Broadcom's Videocore IV GPU, although we did
not use the GPU in our evaluation, due to the lack of sofware support
\footnote{TensorFlow currently only supports NVIDIA's GPUs.}.
For software, we use Ubuntu MATE 16.04, TensorFlow 1.1.
We disabled DVFS (dynamic voltage frequency scaling) and
configured the clock speed of each core statically at the maximum 1.2GHz.

\subsection{Inference Timing for Real-Time Control}

For real-time control of a car (or any robot), the control loop
frequency must be sufficiently high so that the car can quickly
react to the changing environment and its internal states. In general,
control performance improves when the frequency is higher, though
computation time and the type of particular physical system are
factors in determining a proper control loop frequency. While a standard
control system may be comprised of multiple control loops with
differing control frequencies---e.g., an inner control loop for lower-level
PD control, an outer loop for motion planning, etc.---DeepPicar's
control loop is a single layer, as shown earlier in
Figure~\ref{fig:controlloop}, since a single deep neural network
replaces the traditional multi-layer control pipline. (Refer to
Figure~\ref{fig:end-to-end-control} on the differences between the
standard robotics control vs. end-to-end deep learning approach).
This means that the DNN inference operation must be completed
within the inner-most control loop update frequency. To understand
achievable control-loop update frequencies, we experimentally measured
the execution times of DeepPicar's DNN inference operations.

@@

\begin{figure}[h]
  \centering
  \includegraphics[width=.5\textwidth]{figs/Fig7_new}
  \caption{DeepPicar's control loop processing times over 1000 input image frames.}
  \label{fig:control-loop-timing}
\end{figure}

@@

In this experiment, we investigate the scalability of performing
inference operations of DeepPicar's neural network with respect to the
number of cores. As noted earlier, the Raspberry Pi 3 platform has
four Cortex-A53 cores and TensorFlow 
provides a programmable mechanism to adjust how many cores are to be
used by the library. Leveraging this feature, we repeat the
same experiment described in the previous subsection with varying
numbers of CPU cores---from one to four.

@@

Figure~\ref{fig:perf-vs-corecnt} shows the average execution time of
the control loop as we vary the number of cores used by
TensorFlow. As expected, as we assign more cores, the average execution
time decreases---from 46.30 ms on a single core to 22.86 ms on four
cores (over a 100\% improvement). However, the improvement is far from an ideal
linear scaling. In particular, from 3 cores to 4 cores, the
improvement is mere 2.80 ms, or 12\%. In short, we find that the
scalability of DeepPicar's deep neural network is not ideal on the
platform.

As noted in~\cite{NVIDIA2015}, DNN inferencing is inherently more
difficult to parallelize than training because the easiest
parallelization option, batching (i.e., processing multiple images in
parallel), is not available or is limited. Specifically, in DeepPicar,
only one image frame, obtained from the camera, can be processed at a
time. Thus, more fine-grained algorithmic parallelization is needed to
improve inference performance~\cite{NVIDIA2015}, which generally does
not scale very well.

On the other hand, the poor scalability opens up the possibility of
consolidating multiple different tasks or different neural network
models rather than allocating all cores for a single neural network
model.
For example, it is conceivable to use four cameras and four different
neural network models, each of which is trained separately for
different purpose and executed on a single dedicated core.
Assuming we use the same network
architecture for all models, then one might expect to achieve up to
20 Hz using one core (given 1 core can deliver 46 ms average
execution time).
In the next experiment, we investigate the
feasibility of such a scenario.

@@

These increases in inference times in the co-scheduled scenarios are
expected because it is well-known that co-scheduled tasks on a
multicore platform interfere with each other due to contention in the
shared hardware resources, such as the shared
cache~\cite{Gracioli2015} and DRAM~\cite{Yun2013}.

@@

Figure~\ref{fig:perf_vs_bandwidth} shows the execution times of the DNN 
model against varying number of BwRead and BwWrite co-runners. Note
that BwWrite cause significantly high execution time incrase---up 
to 11.6X---on the DNN inferencing timing while BwRead causes
relatively much smaller time increase. While execution time increase
was expected, the degree of which is quite surprising---\emph{if the DNN
controller was driving an actual car, it would result in a crash of the
car!}

Given the importance of predictable timing on real-time control, such
as our DNN based control task, we wanted to understand (1) the main
source of the timing and (2) to evaluate existing isolation methods to
avoid this kind of timing interference. Specifically, shared cache
space and DRAM bandwidth are two most well-known sources of contention
in multicore. Thus, in the following, we investigate if and how
they influence the observed timing interference and how effective are
the existing mitigation methods. To our surprise, we find that neither
the cache space nor DRAM bandwidth is the root cause of the
DeepPicar's observed worst-case execution time increase.

>>>>

\subsection{Setup}
The Raspberry Pi 3 Model B platform used in DeepPicar equips a Broadcom
BCM2837 SoC, which has a quad-core ARM Cortex-A53 cluster,
running at up to 1.2GHz. Each core has private 32K I/D caches
and all cores share a 512KB L2 cache.
The chip also includes Broadcom's Videocore IV GPU, although we did
not use the GPU in our evaluation, due to the lack of sofware support
\footnote{TensorFlow currently only supports NVIDIA's GPUs.}.
For software, we use Ubuntu MATE 16.04, TensorFlow 1.1.
We disabled DVFS (dynamic voltage frequency scaling) and
configured the clock speed of each core statically at the maximum 1.2GHz.
We use the SCHED\_FIFO real-time scheduler to schedule the DNN control
task while using the CFS when executing memory intensive co-runners.

\subsection{Inference Timing for Real-Time Control}
For real-time control of a car (or any robot), the control loop
frequency must be sufficiently high so that the car can quickly
react to the changing environment and its internal states. In general,
control performance improves when the frequency is higher, though
computation time and the type of particular physical system are
factors in determining a proper control loop frequency. While a standard
control system may be comprised of multiple control loops with
differing control frequencies---e.g., an inner control loop for lower-level
PD control, an outer loop for motion planning, etc.---DeepPicar's
control loop is a single layer, as shown earlier in
Figure~\ref{fig:controlloop}, since a single deep neural network
replaces the traditional multi-layer control pipline. (Refer to
Figure~\ref{fig:end-to-end-control} on the differences between the
standard robotics control vs. end-to-end deep learning approach).
This means that the DNN inference operation must be completed
within the inner-most control loop update frequency.

To understand achievable control-loop update frequencies, we
experimentally measured the execution times of DeepPicar's DNN
inference operations.

@@

\begin{figure}[h]
  \centering
  \includegraphics[width=.45\textwidth]{figs/Fig7_new}
  \caption{DeepPicar's control loop processing times over 1000 input image frames.}
  \label{fig:control-loop-timing}
\end{figure}

@@

In this experiment, we investigate the scalability of performing
inference operations of DeepPicar's neural network with respect to the
number of cores. As noted earlier, the Raspberry Pi 3 platform has
four Cortex-A53 cores and TensorFlow 
provides a programmable mechanism to adjust how many cores are to be
used by the library. Leveraging this feature, we repeat the
experiment in the previous subsection but with using varying
numbers of CPU cores---from one to four.

@@

Figure~\ref{fig:perf-vs-corecnt} shows the average execution time of
the control loop as we vary the number of cores used by
TensorFlow. As expected, as we assign more cores, the average execution
time decreases---from 46.30 ms on a single core to 22.86 ms on four
cores (over a 100\% improvement). However, the improvement is far from an ideal
linear scaling. In particular, from 3 cores to 4 cores, the
improvement is mere 2.80 ms, or 12\%. In short, we find that the
scalability of DeepPicar's deep neural network is not ideal.

As noted in~\cite{NVIDIA2015}, DNN inferencing is inherently more
difficult to parallelize than training because the easiest
parallelization option, batching (i.e., processing multiple images in
parallel), is not available or is limited. Specifically, in DeepPicar,
only one image frame, obtained from the camera, can be processed at a
time. Thus, more fine-grained algorithmic parallelization is needed to
improve inference performance~\cite{NVIDIA2015}, which generally does
not scale well.

On the other hand, the limited scalability opens up the possibility of
consolidating multiple different tasks or different neural network
models rather than allocating all cores for a single neural network
model.
For example, it is conceivable to use four cameras and four different
neural network models, each of which is trained separately for
different purpose and executed on a single dedicated core.
Assuming we use the same network
architecture for all models, then one might expect to achieve up to
20 Hz using one core (given 1 core can deliver 46 ms average
execution time).
In the next experiment, we investigate the
feasibility of such a scenario.

@@

These increases in inference times in the co-scheduled scenarios are
expected because co-scheduled tasks on a
multicore platform interfere with each other due to contention in the
shared hardware resources, such as the shared
cache and DRAM~\cite{Gracioli2015,Yun2013}.

@@

Figure~\ref{fig:perf_vs_bandwidth} shows the result. Note
that BwWrite co-runners cause significantly high execution time
incrases---up to 11.6X---on DNN inferencing while BwRead co-runners
cause relatively much smaller time increases. While execution time
increases are expected, the degree of which in the worst-case is quite
surprising---\emph{if the DNN controller was driving an actual car, it
  would result in a crash of the car!}

Given the importance of predictable timing on real-time control, such
as our DNN based control task, we wanted to understand (1) the main
source of the timing and (2) to evaluate existing isolation methods to
avoid this kind of timing interference. Specifically, shared cache
space and DRAM bandwidth are two most well-known sources of contention
in multicore. Thus, in the following, we investigate if and how
they influence the observed timing interference and how effective are
the existing mitigation methods.

%% To our surprise, we find that neither
%% the cache space nor DRAM bandwidth is the root cause of the
%% DeepPicar's observed worst-case execution time increase.

### 2018-04-19 paper/intro.tex
revision

% what are our findings?
Using DeepPicar, we systematically analyze its real-time
capabilities in the context of deep-learning based real-time
control, especially on real-time deep neural network \emph{inferencing}.
We also evaluated other, more powerful, embedded computing
platforms to better understand achievable real-time performance of
DeepPicar's deep-learning based control system and the impact of
computing hardware architectures.

% Findings:
%
% - we find DNN processing is an ideal real-time workload from the
%   perspective of timing predictability as the amount of computation
%   needed is fixed at design time and doesn't change over input.
%
% - real-time processing of non-trivial DNN inferencing is feasible
%   on today's embedded multicore platform, even as inexpensive as
%   raspberry pi3.
%
% - all tested platforms have enough computing capacity to consolidate 
%   multiple DeepPicar's CNN workloads or other compute intensive tasks
%   simultaneously.
%
% - however, when consolidating multiple workloads shared resource
%   contention is an important issue. we carefully analyzed
%   sensitivity of DeepPicar's workload with respect to shared
%   hardware resources.
%
% - in particular, we find DNN processing is highly sensitive to
%   memory performance. performance severely affected when memory
%   bandwidth is contented.
%
% - however, we find that dnn processing is not sensitive to cache
%   space. we find applying cache partitioning does not help improve
%   isolation of DNN workload when memory bandwidth is contended.
From the systematic study, we have made a number of interesting
observations.
% predictability
First, we find that CNN inferencing is highly time predictable as the
amount of computation needed to complete a 
single inference is fixed at the network architecture design time and
does not change at runtime over different inputs (e.g., different image
frames). This predictable timing behavior is obviously a desirable
property for real-time systems.
%% , which gives another reason for the
%% real-time community to be interested in the use of DNNs.

% performance on low-cost embedded multicore
Second, we find that real-time processing of CNN inferencing is
feasible on today's embedded computing platforms, even one   
as inexpensive as the Raspberry Pi 3. For example, just one Cortex-A53
core of the Raspberry Pi3 is able to complete the CNN inferencing task
in less than 50ms, sufficient to support 20 Hz control loop. 
%% We find that the control loop 
%% completes in under 33.$\overline{\mbox{33}}$ ms, or 30 Hz, using just two 
%% cores of the Raspberry Pi 3's quad-core CPU (w/o using its GPU), and can
%% do so 100\% of the time; or we can achieve 20 Hz control frequency using
%% just one core.
%% Other tested embedded platforms, the Intel UP and NVIDIA TX2, offer even
%% better performance, and are capable of supporting deep-learning based 
%% real-time control over 100 Hz control frequency on the TX2 when its
%% embedded GPU is used.
We find that all the tested embedded computing platforms have 
enough computing capacity to consolidate multiple instances of
DeepPicar's DNN workload.

% all platforms have enough capacity to consolidate
Third, contention in shared hardware resources remains an important
issue when consolidating multiple workloads. 
We observe up to 11.6X control
loop execution time increase due to shared resource contention when
memory intensive  applications are co-scheduled on idle CPU cores.
We find that DeepPicar's inferencing task is highly
sensitive to memory bandwidth, but not cache space.

@@

% contributions
The {\bf contributions} of this paper are as follows:
\begin{itemize}
  \item We present DeepPicar, a low-cost autonomous car testbed using
    a state-of-the-art CNN for end-to-end real-time control, to lower
    economic and safety-related barriers to the study autonomous
    cars.~\footnote{The source code, dataset, and build instructions
      of DeepPicar can be found at:\fixme{Michael github url}}

  \item We provide extensive systematic experimental results using
    DeepPicar's CNN inferencing workload on a number of representative
    contemporary embedded computing platforms. The results provide
    insights for real-time processing of the CNN workload and the
    effect of consolidation.

    %% While larger embedded
    %% platforms have been evaluated, as in  \cite{Otterness2017}, to the
    %% best of our knowledge, no prior work investigated the
    %% impacts of shared resource contention on DNN inferencing workloads
    %% in the context of smaller embedded platforms like the Raspberry Pi 3.
\end{itemize}

>>>>

% what are our findings?
Using DeepPicar, we systematically analyze its real-time
capabilities in the context of deep-learning based real-time
control, especially on real-time deep neural network \emph{inferencing}.
We also evaluated other, more powerful, embedded computing
platforms to better understand achievable real-time performance of
DeepPicar's deep-learning based control system and the impact of
computing hardware architectures.
From the systematic study, we want to answer the following questions:
(1) how well does the DNN based real-time inferencing task perform on
contemporary embedded multicore computing platforms? (2) how
susceptible is the DNN inferencing task to contention in shared
hardware resources (e.g., cache and DRAM) when multiple tasks/models
are consolidated? (3) are the existing state-of-the-art shared resource
isolation techniques effective in protecting real-time performance of
the DNN inferencing task?

% Findings:
%
% - we find DNN processing is an ideal real-time workload from the
%   perspective of timing predictability as the amount of computation
%   needed is fixed at design time and doesn't change over input.
%
% - real-time processing of non-trivial DNN inferencing is feasible
%   on today's embedded multicore platform, even as inexpensive as
%   raspberry pi3.
%
% - all tested platforms have enough computing capacity to consolidate 
%   multiple DeepPicar's CNN workloads or other compute intensive tasks
%   simultaneously.
%
% - however, when consolidating multiple workloads shared resource
%   contention is an important issue. we carefully analyzed
%   sensitivity of DeepPicar's workload with respect to shared
%   hardware resources.
%
% - in particular, we find DNN processing is highly sensitive to
%   memory performance. performance severely affected when memory
%   bandwidth is contented.
%
% - however, we find that dnn processing is not sensitive to cache
%   space. we find applying cache partitioning does not help improve
%   isolation of DNN workload when memory bandwidth is contended.
Our observations are as follows.
First, we find that real-time processing of the DNN inferencing is
feasible on contemporary embedded computing platforms, even one   
as inexpensive as the Raspberry Pi 3. Second, while consolidating
multiple DNN models and task are feasible but the impact of shared
resource contention can be catastrophic in the worst-case as we
observe up to 11.6X slowdown even when the DNN model was given a
dedicated core. Third, using the state-of-the-art OS-level cache
partitioning and memory bandwidth control
mechanisms~\cite{yun2014rtas,Yun2013}, we find that cache space has a
minimal performance impact to the DNN workload while memory
performance has a significant impact.

%% % predictability
%% First, we find that CNN inferencing is highly time predictable as the
%% amount of computation needed to complete a 
%% single inference is fixed at the network architecture design time and
%% does not change at runtime over different inputs (e.g., different image
%% frames). This predictable timing behavior is obviously a desirable
%% property for real-time systems.
%% %% , which gives another reason for the
%% %% real-time community to be interested in the use of DNNs.
% performance on low-cost embedded multicore
%% For example, just one Cortex-A53
%% core of the Raspberry Pi3 is able to complete the CNN inferencing task
%% in less than 50ms, sufficient to support 20 Hz control loop. 
%% We find that the control loop 
%% completes in under 33.$\overline{\mbox{33}}$ ms, or 30 Hz, using just two 
%% cores of the Raspberry Pi 3's quad-core CPU (w/o using its GPU), and can
%% do so 100\% of the time; or we can achieve 20 Hz control frequency using
%% just one core.
%% Other tested embedded platforms, the Intel UP and NVIDIA TX2, offer even
%% better performance, and are capable of supporting deep-learning based 
%% real-time control over 100 Hz control frequency on the TX2 when its
%% embedded GPU is used.
%% We find that all the tested embedded computing platforms have 
%% enough computing capacity to consolidate multiple instances of
%% DeepPicar's DNN workload.

% all platforms have enough capacity to consolidate
%% Third, contention in shared hardware resources remains an important
%% issue when consolidating multiple workloads. 
%% We observe up to 11.6X control
%% loop execution time increase due to shared resource contention when
%% memory intensive  applications are co-scheduled on idle CPU cores.
%% We find that DeepPicar's inferencing task is highly
%% sensitive to memory bandwidth, but not cache space.

@@

% contributions
This paper makes the following {\bf contributions}:
\begin{itemize}
  \item We present DeepPicar, a low-cost autonomous car testbed, which
    employs a state-of-the-art CNN based end-to-end
    real-time control.~\footnote{The source code, datasets, and build
      instructions of DeepPicar can be found at:
      \url{https://github.com/mbechtel2/DeepPicar-v2}}.
  \item We provide extensive empirical performance evaluation results
    of the CNN inferencing workload on a number of contemporary
    embedded computing platforms.
  \item We apply the state-of-the-art shared resource isolation
    techniques and evaluate their effectiveness in protecting the
    inferencing workload in consolidated settings.
\end{itemize}

### 2018-04-19 paper/overview.tex
remove fixme. replace track 1 picture.
revision

@@

In this section, we provide an overview of our DeepPicar platform.
In developing DeepPicar, one of our primary goals is to faithfully
replicate NVIDIA's DAVE-2 system on a smaller scale using a low cost
multicore platform, the Raspberry Pi 3. Because the Raspberry Pi 3's
computing performance is much lower than that of the DRIVE PX platform
used in DAVE-2, we are interested in if, and how, we can process
computationally expensive neural network operations in
real-time. Specifically, inferencing (forward pass processing)
operations must be completed within each control period
duration---e.g., a WCET of 33.$\overline{\mbox{33}}$ ms for 30Hz control 
frequency---locally on the Pi 3 platform, although training of the 
network (back-propagation for weight updates) can be done offline and 
remotely using a desktop computer.
Figure~\ref{fig:overview} shows the DeepPicar, which is comprised of a
set of inexpensive components: a Raspberry Pi 3 Single Board Computer
(SBC), a Pololu DRV8835 motor driver, a Playstation Eye webcam, a
battery, and a 1:24 scale RC car. Table~\ref{tbl:carbom} shows the
estimated cost of the system.

@@

For the neural network architecture, we implement
NVIDIA DAVE-2's convolutional neural network (CNN) using a TensorFlow
based CNN model in ~\cite{deeptesla}. Note, however, that the CNN model
in~\cite{deeptesla} was considerably larger than NVIDIA's CNN model as
it contained an additional fully-connected layer of approximately 1.3M
additional parameters. We eliminated the additional layer to
faithfully recreate the NVIDIA's original CNN model.
%% The main difference is that we do not utilize a normalization layer, and 
%% instead initialize the weights using a Xavier initialization. 
As in DAVE-2, the CNN takes a raw color image (200x66 RGB pixels)
as input and produces a single steering angle value as
output. Figure~\ref{fig:architecture} shows the network architecture
used in this paper, which is comprised of 9 layers, 250K parameters,
and about 27 million connections as in NVIDIA DAVE-2's.
%% Note, however,
%% that we did not apply the normalization mentioned
%% in~\cite{Bojarski2016}, as it does not include trainable parameters
%% and its computational demand with respect to the overall CNN
%% processing is minimal.

\begin{figure}[h]
  \centering
  \begin{subfigure}{0.4\textwidth}
    \includegraphics[width=\textwidth]{figs/track_new2}
    \caption{Track 1.}
    \label{fig:track}
  \end{subfigure}
  \hfill
  \begin{subfigure}{0.4\textwidth}
    \includegraphics[width=\textwidth]{figs/track-3}
    \caption{Track 2.}
    \label{fig:track2}
  \end{subfigure}
  \caption{Custom tracks used for training/testing}
  \label{fig:tracks}
\end{figure}

@@

% inferencing on pi3
Once the network is trained on the desktop computer, the trained model
is copied back to the Raspberry Pi 3. The network is then used
by the car's main controlller, which feeds an image frame from the web
camera as input to the network at each control period. The network
produced steering angle output is then converted into the PWM values
of the car's steering motor. Figure~\ref{fig:controlloop} shows simplified 
pseudo code of the controller's main loop. Among the five steps, the 3rd step, 
network inferencing, is the most computationally intensive and dominates the
execution time.

Note that although the steering angle output of the network ($angle$) is
a continuous real value, the RC car platform we used unfortunately
only supports three discrete angles---left (-30$^{\circ}$), center
(0$^{\circ}$), and right (+30$^{\circ}$)---as control inputs.
Currently, we approximate the network generated real-valued
angle to the closest one of the three angles, which may
introduce inaccuracy in control.
%% In the future, we plan to use a different (more expensive) RC car
%% platform that can precisely control the car's steering angle.

Other issues we observed in training/testing the network, which
can affect the prediction accuracy of the CNN, are camera and actuator
(motor) control latencies. In DeepPicar's
context, the camera latency is from the time the camera
sensor observes the scene to the time the computer actually reads the
digitized image data. This time can be noticable depending on the camera 
used and the performance of the Pi. Higher camera latency could
negatively affect control performance, because the DNN would analyze
stale scenes. Actuator (motor) control latency also can be a
factor, because moving the steering motor to a desired position takes
time. 
In our platform, the combined latency is measured to be around
50 ms. In other words, in the recorded video and control data, a
control action (left, center, right) taken by the CNN appears to be
applied 50 ms later. Considering that camera's framerate is 30Hz
(33.3 ms/frame), this is about two frames after the control action.
%% We experimentally measured the camera
%% latency and found it to be around 50-100 ms.
If this value is too high, control performance may suffer.
Our initial prototype suffered this problem as the observed camera and
control latency was as high as 300 ms, which negatively affected
control performance.
For reference, the latency of human perception is known to be as fast
as 13 ms~\cite{ThomasBurger2015}. 
% https://www.pubnub.com/blog/2015-02-09-how-fast-is-realtime-human-perception-and-technology/

Despite these issues, the trained models were still able to
achieve a reasonable degree of accuracy, successfully navigating
several different tracks we used for training. In our testing, we found 
that the DeepPicar could remain on a fairly challenging track for over 
10 minutes at a moderate speed (50\% throttle), at which point we stopped
the experiment, and more than one minute at a higher speed (75\%
throttle)~\footnote{A collection of self-driving videos can be found
  at: \fixme{Michael: link your video}
  %% \url{https://photos.app.goo.gl/ce93sU7jPk4ywO8u2}
}. Running at
higher speed is inherently more challenging because the CNN controller
has less time to recover from mistakes (bad predictions).  Also, we
find that the prediction accuracy is significantly affected by the
quality of training data as well as various environmental factors such
as lighting conditions. We plan to investigate more systematic ways
to improve the CNN's prediction accuracies.

>>>>

In this section, we provide an overview of our DeepPicar platform.
In developing DeepPicar, one of our primary goals is to faithfully
replicate NVIDIA's DAVE-2 system on a smaller scale using a low cost
multicore platform, the Raspberry Pi 3. Because Raspberry Pi 3's
computing performance is much lower than that of the DRIVE
PX~\cite{drivepx} platform used in DAVE-2, we are interested in if,
and how, we can process 
computationally expensive neural network operations in
real-time. Specifically, inferencing (forward pass processing)
operations must be completed within each control period
duration---e.g., a WCET of 33.3 ms for 30Hz control 
frequency---locally on the Pi 3 platform, although training of the 
network (back-propagation for weight updates) can be done offline and 
remotely using a desktop computer.

@@

Figure~\ref{fig:overview} shows the DeepPicar, which is comprised of a
set of inexpensive components: a Raspberry Pi 3 Single Board Computer
(SBC), a Pololu DRV8835 motor driver, a Playstation Eye webcam, a
battery, and a 1:24 scale RC car. Table~\ref{tbl:carbom} shows the
estimated cost of the system.

For the neural network architecture, we implement NVIDIA DAVE-2's
convolutional neural network (CNN) using an open-source CNN model in
~\cite{deeptesla}. Note, however, that the CNN model
in~\cite{deeptesla} is considerably larger than NVIDIA's CNN
model as it contains an additional fully-connected layer of
approximately 1.3M additional parameters. We remove the additional
layer to faithfully recreate NVIDIA's original CNN model.
%% The main difference is that we do not utilize a normalization layer, and 
%% instead initialize the weights using a Xavier initialization. 
As in DAVE-2, the CNN takes a raw color image (200x66 RGB pixels)
as input and produces a single steering angle value as output.
Figure~\ref{fig:architecture} shows the network architecture
used in this paper, which is comprised of 9 layers, 250K parameters,
and about 27 million connections as in NVIDIA DAVE-2's architecture.

@@

%% Note, however,
%% that we did not apply the normalization mentioned
%% in~\cite{Bojarski2016}, as it does not include trainable parameters
%% and its computational demand with respect to the overall CNN
%% processing is minimal.

\begin{figure}[h]
  \centering
  \begin{subfigure}{0.4\textwidth}
    \includegraphics[width=\textwidth]{figs/track-2}
    \caption{Track 1.}
    \label{fig:track}
  \end{subfigure}
  \hfill
  \begin{subfigure}{0.4\textwidth}
    \includegraphics[width=\textwidth]{figs/track-3}
    \caption{Track 2.}
    \label{fig:track2}
  \end{subfigure}
  \caption{Custom tracks used for training/testing}
  \label{fig:tracks}
\end{figure}

@@

% inferencing on pi3
Once the network is trained on the desktop computer, the trained model
is copied back to the Raspberry Pi 3. The network is then used
by the car's main controlller, which feeds image frames from the web
camera as input to the network in real-time. At each control period,
the network produced steering angle output is converted into the PWM values
of the car's steering motor. Figure~\ref{fig:controlloop} shows simplified 
pseudo code of the controller's main loop. Among the five steps, the 3rd step, 
network inferencing, is the most computationally intensive and dominates the
execution time.

Note that although the steering angle output of the network ($angle$) is
a continuous real value, the RC car platform we used only supports
three discrete angles---left (-30$^{\circ}$), center 
(0$^{\circ}$), and right (+30$^{\circ}$)---as control inputs.
We approximate the network generated real-valued angle to the closest
one of the three angles. Although this may introduce inaccuracy in
control, the observed control performance of the system is respectable,
likely due to the inherently stocastic nature of DNN.

%% In the future, we plan to use a different (more expensive) RC car
%% platform that can precisely control the car's steering angle.

Another factors that can potentially affect the prediction accuracy of
the CNN, are camera and actuator (motor) control latencies. The camera
latency is from the time the camera sensor observes the scene to the
time the computer actually reads the digitized image data. This time
can be noticable depending on the camera used and the data processing
time of the computing platform. Higher camera latency could
negatively affect control performance, because the DNN would analyze
stale scenes. The actuator (motor) control latency from the time
the control output is sent to the steering motor to the time the motor
actually moved at a desired position, which also can takes
considerable time. In our platform, the combined latency is measured
to be around 50 ms, which is reasonble.
%% In other words, %% in the recorded video and control data, 
%% a control action of the CNN appears to be applied 50 ms later.
%% Considering that camera's framerate is 30Hz
%% (33.3 ms/frame), this is about two frames after the control action.
%% We experimentally measured the camera
%% latency and found it to be around 50-100 ms.
If this value is too high, control performance may suffer.
Our initial prototype suffered this problem as the observed latency
was as high as 300 ms, which negatively affected control performance.
For reference, the latency of human perception is known to be as fast
as 13 ms~\cite{ThomasBurger2015}. 
% https://www.pubnub.com/blog/2015-02-09-how-fast-is-realtime-human-perception-and-technology/

Our trained DNN models showed good prediction accuracy, successfully
navigating several different tracks we used for training.
For instance, the DeepPicar could remain on Track 1
(Figure \label{fig:track}) over 10 minutes at a moderate speed (50\%
throttle), at which point we stopped the experiment, and more than one
minute at a higher speed (75\% throttle)~\footnote{Self-driving videos: \url{https://photos.app.goo.gl/q40QFieD5iI9yXU42}
  %% \url{https://photos.app.goo.gl/ce93sU7jPk4ywO8u2}
}. Running at
higher speed is inherently more challenging because the CNN controller
has less time to recover from mistakes (bad predictions).  Also, we
find that the prediction accuracy is significantly affected by the
quality of training data as well as various environmental factors such
as lighting conditions. We plan to investigate more systematic ways
to improve the CNN's prediction accuracies.
